Static Analysis by Path Focusing

### 2011-06-13 report/report.tex
report += suggestions de matthieu

\lstnewenvironment{Csmall}
{\lstset{language=C,
		basicstyle=\ttfamily\small,
		commentstyle=\color{moka}\textit,
		keywordstyle=\color{minuit},
		identifierstyle=\color{trefle},
		showstringspaces=false}}
{}
\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aimed at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
%\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyser, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of these two methods.
\end{itemize}

@@

  \path [transition] 
		(n0) edge              node {$x = 0$} (n1);
  \path [transition] 
        (n1) edge			   node [above] {$x \geq 100?$} (n3);
  \path [transition] 
        (n1) edge			   node [left] {$x < 100?$} (n2);
  \path [transition] 
        (n2) edge [out=0, in=0, distance=2cm] node [right] {$x = x+1$} (n1);

@@

\emph{Path Focusing} appears to perform similarly or better than 
\emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
containing control-flow merges inside of loops, thanks to the distinction
between all the paths of the loop. In simpler programs, it performs as well as
\emph{Lookahead Widening}

@@

Boustr.c & 0.100s & 0.088s & \begin{tabular}{l}
 -2x.0 + d.0 + 1999 >= 0 \\
 -2x.0 + 3d.0 + 2001 >= 0 \\
 -d.0 + 1 >= 0
\end{tabular}& $\subset$ & 0.064 & \begin{tabular}{l}
 -1996x.0 + 4991d.0 + 1998995 >= 0\\
 -2x.0 + d.0 + 1999 >= 0 \\
 -d.0 + 1 >= 0 \\
\end{tabular} \\ \hline

ex9.cpp & 0.084s & 0.082s & \begin{tabular}{l}
 r.0 = 0 \\
-i.0 + 100 >= 0 \\
i.0 >= 0
\end{tabular}& $\subset$ & 0.052 & \begin{tabular}{l}
-i.0 + 100 >= 0\\
r.0 >= 0 \\
i.0 - r.0 >= 0 \\
24i.0 - 25r.0 + 75 >= 0
\end{tabular} \\ \hline

@@

ex6.c & 0.104s & $\infty$ & \begin{tabular}{l}
-y.0 + 51 >= 0 \\
y.0 >= 0 \\
x.0 - 51y.0 + 2550 >= 0 \\
x.0 >= 0
\end{tabular}& $\subset$ & 0.032 & \begin{tabular}{l}
y.0 >= 0
\end{tabular} \\ \hline

ex8.c & 0.516s & 0.408s & \begin{tabular}{l}
 -y.0 - x.0 + 1 >= 0\\
 -y.0 + x.0 + 1 >= 0\\
 y.0 - x.0 + 1 >= 0\\
 y.0 + x.0 + 1 >= 0
\end{tabular}& $\subset$ & 0.208 & \begin{tabular}{l}
-x.0 + 1 >= 0 \\
x.0 + 1 >= 0 
\end{tabular} \\ \hline

ex8c.c & 0.656s & 0.548s & \begin{tabular}{l}
c.0 + a.0 = 0\\
b.0 + d.0 = 0\\
-b.0 - c.0 + 1 >= 0\\
-b.0 + c.0 + 1 >= 0\\
b.0 - c.0 + 1 >= 0\\
b.0 + c.0 + 1 >= 0\\
\end{tabular}& $\subset$ & 0.916s & \begin{tabular}{l}
-b.0 + 1 >= 0\\
-b.0 + a.0 + 1 >= 0\\
-c.0 + 1 >= 0\\
-c.0 + d.0 + 1 >= 0\\
-a.0 + 1 >= 0\\
-d.0 + 1 >= 0\\
a.0 + 1 >= 0\\
c.0 - d.0 + 1 >= 0\\
c.0 + d.0 + 1 >= 0\\
b.0 - a.0 + 1 >= 0\\
b.0 + 1 >= 0\\
\end{tabular} \\ \hline

ex8b.c & 1.016s & 0.816s & \begin{tabular}{l}
c.0 + y.0 = 0\\
c.0 + a.0 = 0\\
b.0 + d.0 = 0\\
b.0 + x.0 = 0\\
-b.0 - c.0 + 1 >= 0\\
-b.0 + c.0 + 1 >= 0\\
b.0 - c.0 + 1 >= 0\\
b.0 + c.0 + 1 >= 0\\
\end{tabular}& $\subset$ & 36.30s & \begin{tabular}{l}
-b.0 + 1 >= 0 \\
-b.0 + a.0 + 1 >= 0\\
-c.0 + 1 >= 0\\
-c.0 + d.0 + 1 >= 0\\
-a.0 + 1 >= 0\\
-y.0 - x.0 + 1 >= 0\\
-y.0 + x.0 + 1 >= 0\\
-d.0 + 1 >= 0\\
x.0 + 1 >= 0\\
y.0 - x.0 + 1 >= 0\\
y.0 + 1 >= 0\\
a.0 + 1 >= 0\\
c.0 - d.0 + 1 >= 0\\
c.0 + d.0 + 1 >= 0\\
b.0 - a.0 + 1 >= 0\\
b.0 + 1 >= 0\\
\end{tabular}\\ \hline

@@

\begin{itemize}
\item We use the domain of convex polyhedra as the abstract domain. Since this
domain requires a widening operator, a more subtle analysis technique may
sometimes discover invariants that are less precise than a more simplified
technique.
\item Our technique performs well in the case of self loops in the multigraph,
thanks to the widening/narrowing phases (or acceleration). Yet, in the case of
paths that do not constitute a self loop in the multigraph, one perform a
widening operation, but no narrowing iteration. Apparently, lots of the cases
where \emph{Lookahead Widening} performs better than \emph{Path Focusing} are
due to this drawback. An example of such loss of precision is depicted in
appendix \ref{lostprecision}.
An good idea that could solve this problem is to try to combine
the two methods.
\end{itemize}

@@

	\section{Combining Lookahead Widening and Path Focusing}

	Path Focusing technique works on the expanded multigraph.
	Its limitation is that it does narrowing
	sequences during the iterations only for self-loops of this multigraph.

@@

	Our analyser is able to deal with small- and middle-sized programs, but
	looses
	precision in the case of programs with pointers or several functions.
	It has been tested with program having loops containing several hundreds
	lines of code, and gave results after a time in the order of the second.
	It could be improved in many ways, in order to support
	for instance interprocedural analysis. A memory model could be added to
	handle programs with pointers and \emph{load/store} instructions.
	One could also try to compute disjunctive invariants using the technique
	described in this report.
	Since our technique combines well with acceleration techniques, another
	direction is to implement this last technique to improve our results
	obtained with classical widening/narrowing.

>>>>

\lstnewenvironment{Csmall}
{\lstset{language=C,
		basicstyle=\ttfamily\small,
		commentstyle=\color{moka}\textit,
		keywordstyle=\color{minuit},
		identifierstyle=\color{trefle},
		showstringspaces=false}}
{}
\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aimed at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce rapport s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
%\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyser, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of these two methods. This
implementation has been tested on a wide range of programs, including small
test-cases as well as real-life programs.
\end{itemize}

@@

  \path [transition] 
		(n0) edge              node {$x \gets 0$} (n1);
  \path [transition] 
        (n1) edge			   node [above] {$x \geq 100$} (n3);
  \path [transition] 
        (n1) edge			   node [left] {$x < 100$} (n2);
  \path [transition] 
        (n2) edge [out=0, in=0, distance=2cm] node [right] {$x \gets x+1$} (n1);

@@

\emph{Path Focusing} appears to perform similarly or better than 
\emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
containing control-flow merges inside  loops, thanks to the distinction
between all the paths of the loop. In simpler programs, it performs as well as
\emph{Lookahead Widening}

@@

Boustr.c & 0.100s & 0.088s & \begin{tabular}{l}
 -2x.0 + d.0 + 1999 >= 0 \\
 -2x.0 + 3d.0 + 2001 >= 0 \\
 -d.0 + 1 >= 0
\end{tabular}& $\varsubsetneq$ & 0.064 & \begin{tabular}{l}
 -1996x.0 + 4991d.0 + 1998995 >= 0\\
 -2x.0 + d.0 + 1999 >= 0 \\
 -d.0 + 1 >= 0 \\
\end{tabular} \\ \hline

ex9.cpp & 0.084s & 0.082s & \begin{tabular}{l}
 r.0 = 0 \\
-i.0 + 100 >= 0 \\
i.0 >= 0
\end{tabular}& $\varsubsetneq$ & 0.052 & \begin{tabular}{l}
-i.0 + 100 >= 0\\
r.0 >= 0 \\
i.0 - r.0 >= 0 \\
24i.0 - 25r.0 + 75 >= 0
\end{tabular} \\ \hline

@@

ex6.c & 0.104s & $\infty$ & \begin{tabular}{l}
-y.0 + 51 >= 0 \\
y.0 >= 0 \\
x.0 - 51y.0 + 2550 >= 0 \\
x.0 >= 0
\end{tabular}& $\varsubsetneq$ & 0.032 & \begin{tabular}{l}
y.0 >= 0
\end{tabular} \\ \hline

ex8.c & 0.516s & 0.408s & \begin{tabular}{l}
 -y.0 - x.0 + 1 >= 0\\
 -y.0 + x.0 + 1 >= 0\\
 y.0 - x.0 + 1 >= 0\\
 y.0 + x.0 + 1 >= 0
\end{tabular}& $\varsubsetneq$ & 0.208 & \begin{tabular}{l}
-x.0 + 1 >= 0 \\
x.0 + 1 >= 0 
\end{tabular} \\ \hline

ex8c.c & 0.656s & 0.548s & \begin{tabular}{l}
c.0 + a.0 = 0\\
b.0 + d.0 = 0\\
-b.0 - c.0 + 1 >= 0\\
-b.0 + c.0 + 1 >= 0\\
b.0 - c.0 + 1 >= 0\\
b.0 + c.0 + 1 >= 0\\
\end{tabular}& $\varsubsetneq$ & 0.916s & \begin{tabular}{l}
-b.0 + 1 >= 0\\
-b.0 + a.0 + 1 >= 0\\
-c.0 + 1 >= 0\\
-c.0 + d.0 + 1 >= 0\\
-a.0 + 1 >= 0\\
-d.0 + 1 >= 0\\
a.0 + 1 >= 0\\
c.0 - d.0 + 1 >= 0\\
c.0 + d.0 + 1 >= 0\\
b.0 - a.0 + 1 >= 0\\
b.0 + 1 >= 0\\
\end{tabular} \\ \hline

ex8b.c & 1.016s & 0.816s & \begin{tabular}{l}
c.0 + y.0 = 0\\
c.0 + a.0 = 0\\
b.0 + d.0 = 0\\
b.0 + x.0 = 0\\
-b.0 - c.0 + 1 >= 0\\
-b.0 + c.0 + 1 >= 0\\
b.0 - c.0 + 1 >= 0\\
b.0 + c.0 + 1 >= 0\\
\end{tabular}& $\varsubsetneq$ & 36.30s & \begin{tabular}{l}
-b.0 + 1 >= 0 \\
-b.0 + a.0 + 1 >= 0\\
-c.0 + 1 >= 0\\
-c.0 + d.0 + 1 >= 0\\
-a.0 + 1 >= 0\\
-y.0 - x.0 + 1 >= 0\\
-y.0 + x.0 + 1 >= 0\\
-d.0 + 1 >= 0\\
x.0 + 1 >= 0\\
y.0 - x.0 + 1 >= 0\\
y.0 + 1 >= 0\\
a.0 + 1 >= 0\\
c.0 - d.0 + 1 >= 0\\
c.0 + d.0 + 1 >= 0\\
b.0 - a.0 + 1 >= 0\\
b.0 + 1 >= 0\\
\end{tabular}\\ \hline

@@

\begin{itemize}
\item We use the domain of convex polyhedra as the abstract domain. Since this
domain requires a widening operator, a more subtle analysis technique may
sometimes discover invariants that are less precise than a more simplified
technique.
\item Our technique performs well in the case of self loops in the multigraph,
thanks to the widening/narrowing phases (or acceleration). Yet, in the case of
paths that do not constitute a self loop in the multigraph, one performs a
widening operation, but no narrowing iteration. Apparently, lots of the cases
where \emph{Lookahead Widening} performs better than \emph{Path Focusing} are
due to this drawback. An example of such loss of precision is depicted in
appendix \ref{lostprecision}.
An idea that could solve this problem is to try to combine
the two methods (see \ref{combining}).
\end{itemize}

@@

	\section{Combining Lookahead Widening and Path Focusing}
	\label{combining}
	Path Focusing technique works on the expanded multigraph.
	Its limitation is that it does narrowing
	sequences during the iterations only for self-loops of this multigraph.

@@

	Our analyser is able to deal with small- and middle-sized test-cases, as
	well as real-life programs. It looses
	precision in the case of programs with pointers or several functions.
	It has been tested with program having loops containing several hundreds
	lines of code, and gave results after a time in the order of the second.
	It could be improved in many ways, in order to support
	for instance interprocedural analysis. A memory model could be added to
	handle programs with pointers and \emph{load/store} instructions.
	One could also try to compute disjunctive invariants using the technique
	described in this report.
	Since our technique combines well with acceleration techniques, another
	direction is to implement this last technique to improve our results
	obtained with classical widening/narrowing.
